Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-08 text/ms.tex
Edit, edit, edit

There is compelling anecdotal evidence for ecological black swans, but systematic evidence has been elusive. A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and a number of major environmental events on geologic and contemporary time scales could be considered to have caused ecological black swans \citep{nunez2012}. Black swans, or heavy-tailed distributions, have been discovered in time-series of marine microbe populations \citet{segura2013} and time to extinction for experimental waterflea populations \citet{drake2014}. As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which found little clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.
% TODO: pull anecdote from that Nunez study

@@

We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity. In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}). We investigated the sensitivity of our results to weaker and stronger priors (Fig.~\ref{fig:priors}, Supporting Information). We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).

@@

We investigated possible covariates of heavy-tailed population dynamics visually and through multilevel modelling. We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu < 10) > 0.5$. We formally investigated these relationships by fitting beta regression multilevel models TODO REF AND MORE DETAILS?. For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means. To account for broad patterns of phylogenetic relatedness, we allowed for varying intercepts at the taxonomic class, order, and species level (Supporting Information).

Finally, we investigated a sample of populations that our method categorized as having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population. These exemplar populations were chosen haphazardly with the purpose of generating hypotheses about the causes of population dynamic black swans.

@@

The majority of our heavy-tailed estimates were robust to alternative population models and observation error (Fig.~\ref{fig:alt}). Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic. However, setting observation error standard deviation to $0.2$ increased the median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt}).

TODO sensitivity to priors Fig.~\ref{fig:alt-priors}.

TODO ADD TO THIS Our simulation testing show that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}) AND ``(as was also observed in simulation testing, Fig.~\ref{fig:sim-nu})''

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a). Longer time-series length was the strongest covariate of observing black-swan dynamics. For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps. However, the absolute change in probability with increased time series length was small ($0.\pHeavyNSixty$ vs.\ $0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

% "Perhaps then you could say "we examined all time series with published explanations of why the black swan events occurred"
TODO STATE HOW THESE 7 WERE CHOSEN
TODO QUANTIFY UPS AND DOWNS --- SKEW
We assembled the documented causes of black-swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table~\ref{tab:sparks}). The majority of documented events were downward black swans and involved a series of unfortunate events. For example, a shortage of nest sites for the shag long-necked bird (TODO SCI.\ NAME) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black swan for a water vole population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.

@@

@@

Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by black-swan events.
TODO END ON THIS

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions. Ecological resource management can draw from other disciplines that focus on heavy tails. For example, earthquake preparedness and response is focussed on black-swan events. Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans (e.g.~REFERENCE?). A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For example, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}. In recent decades ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.

%TODO WORK IN Gerber L.R. & Hilborn R. (2001). Catastrophic events and recovery from low densities in populations of otariids: implications for risk of extinction. Mammal Rev., 31, 131-150.

>>>>

There is compelling anecdotal evidence for ecological black swans, but systematic evidence has been elusive. A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and events such as the global invasion of  Argentine ants and the mutation of viruses to infect new hosts could be considered black swans \citep{nunez2012}. In fact, anecdotes of population catastrophes are numerous and catastrophes may be the most important element affecting population persistence \citep{mangel1994}. For marine mammal populations, we have compelling evidence of catastrophes \citep{gerber2001, ward2007}, and recently, heavy-tailed distributions have been discovered in time-series of marine microbe abundance \citep{segura2013} and time to extinction for experimental waterflea populations \citep{drake2014}. As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which uncovered little clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.
% TODO: pull anecdote from that Nunez study

% TODO \citep{mangel1994} catastrophes may be the most important element affecting species persistence times and more common than we think
% TODO \citep{gerber2001} and \citep{ward2007} evidence for marine mammal populations

@@

We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity. In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}). We investigated the sensitivity of our results to weaker and stronger priors (exponential rate parameter $= 0.005, 0.02$; Fig.~\ref{fig:priors}, Supporting Information). We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).

@@

We investigated possible covariates of heavy-tailed population dynamics visually and through multilevel modelling. We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu < 10) > 0.5$. We formally investigated these relationships by fitting beta regression multilevel models \citep{ferrari2004}. For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means. To account for broad patterns of phylogenetic relatedness, we allowed for varying intercepts (``random effects'') at the taxonomic class, order, and species level (Supporting Information). We fit our model in Stan with weakly informative priors on the coefficients \citep{gelman2008d} and variance parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c} \citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).

Finally, we investigated a sample of populations that our method categorized as having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population. 
%These exemplar populations were chosen haphazardly with the purpose of generating hypotheses about the causes of population dynamic black swans.

@@

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice or priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}). However, setting observation error standard deviation to $0.2$ increased the median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt}). The strength of the prior on $\nu$ had little influence on estimates of black-swan dynamics (Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a). Longer time-series length was the strongest covariate of observing black-swan dynamics. For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps. However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We examined all time series with published explanations of why the black-swan events occurred (Table~\ref{tab:causes-supp}) and focus on seven exemplar populations with high probabilities of heavy tails (Table~\ref{tab:sparks}). The majority (TODO QUANTIFY) of documented events were downward black swans and involved a series of unfortunate events. For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black-swan event for a water vole (\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.

(TODO: substitute a parasite example.)

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions. Ecological resource management can draw from other disciplines that focus on heavy tails. For example, earthquake preparedness and response is focussed on black-swan events. Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans (e.g.~REFERENCE?). A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For example, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}.

@@

Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? 
In recent decades ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.
Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.

### 2014-11-08 text/supp.tex
Edit, edit, edit

TODO

Fig.~\ref{fig:priors}

Fig.~\ref{fig:alt-priors}

@@

We considered a version of the Gompertz model in which an autoregressive
parameter was fit to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align*}
In addition to the parameters in the original Gompertz model, we estimate an additional parameter $\phi$, which represents the dependence of subsequent residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. The MCMC chains did not converge for \modelsNoConvergeAROne\ populations according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains. This included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr($\nu < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model. We did not include these models in Fig.~\ref{fig:alt}. \textit{TODO perhaps I should go up to 16,000 so that more converge.}

@@

TODO, rework this paragraph:
Our results show that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows). Likewise, the other Gompertz parameters were estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols). The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence with observation error is a known phenomenon \citep{knape2012}. Fitting a model with correctly specified observation error made marginal improvements to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).

@@

\singlespacing
\begin{footnotesize}
\begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.3cm}}
\caption{Populations with Pr$(\nu < 10) > 0.5$. TODO}\\
\toprule
\input{../analysis/cause-table.tex}
\label{tab:causes-supp}
\end{longtable}
\end{footnotesize}
\onehalfspacing

@@

% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-3.pdf}
% \caption{
% TODO WILL SPLIT OVER PAGES BY CLASS
%   All filtered times series used in our analysis. The abundances are shown on
%   a log10 vertical axis. Colours indicate taxonomic classes. Open black circles
%   indicate abundance values that were interpolated. Closed black circles
%   indicate single abundance values that were recorded as zero but were set to
%   the next lowest observed abundance.
% }
%     \label{fig:all-ts}
% \end{center}
% \end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/priors-gomp-base.pdf}
\caption{
Probability density of the Bayesian priors for the Gompertz models. From left to right and then top to bottom: (1) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$. The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$. The inset panel shows the same data but with a log-transformed x axis.
}
\label{fig:priors}
\end{center}
\end{figure}

@@

% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
% \caption{
%   Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
%   Gompertz population model. Panels are ordered by increasing Pr($\nu < 10$).
%   Vertical axes are on a log10 scale. Colours indicate taxonomic classes. The
%   labels on each panel indicate Pr($\nu < 10$) $>$ 0.5, the common name for the
%   species, and the Global Population Dynamics Database ID number.
% }
% \label{fig:heavy-ts}
% \end{center}
% \end{figure}
%
% \clearpage

@@

% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
% \caption{
% Simulation testing the Gompertz estimation model when the process deviation
% draws were chosen so that $\nu$ could be estimated close to the true value
% outside the full population model (``effective $\nu$'' within a CV of 0.2 of
% specified $\nu$). The simulation was run across population $\nu$ values
% (columns) and different scenarios (colours): (1) 100 time steps and no
% observation error, (2) 50 time steps and no observation error, (3) 50 time
% steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
% ignored, and (4) 50 time steps with observation error in which the quantity
% of observation error was assumed known. Dashed horizontal lines show the true
% population values; these true values were chosen to represent approximately
% the median values as estimated from the Global Population Dynamics Database.
% Individual dots and lines
% represent a stochastic draws from the true population distribution and
% a model fitting (median and 80\% credible intervals). The panels from top to
% bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
% parameter $\lambda$, and the density dependence parameter $b$.
% }
% \label{fig:sim-gompertz}
% \end{center}
% \end{figure}

% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
% \caption{
%   The distribution of median estimates from the same simulation as described in
%   Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
%   whiskers extend to $1.5$ times the interquartile range and outliers are shown
%   as dots. Note that these distributions represent the output from only 20
%   simulation iterations, so there is some level of remaining sampling noise.
% }
% \label{fig:sim-gompertz-boxplots}
% \end{center}
% \end{figure}

>>>>

To test if the prior on $\nu$ influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors. Our base model used a prior on $\nu$ of Truncated-Exponential(0.02, min.\ = 2). For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}). Note that the base and weaker priors are relatively flat within the region of $\nu < 20$---the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed.

Our results show that these weaker and stronger priors would have little influence on our conclusions about heavy-tailed dynamics (Fig.~\ref{fig:alt-priors}). When the data are informative about tail behaviour (i.e.\ when there is strong evidence of low $\nu$ values, upper-right of Fig.~\ref{fig:alt-priors}), the prior has little impact on the estimate of $\nu$. When the data are less informative about $\nu$ (i.e.\ when there are no or few tail events and time series are short or noisy), the prior can pull the estimate of $\nu$ towards larger or smaller values (Fig.~\ref{fig:alt-priors}). The vast majority of the populations with Pr$(\nu < 10)$ in the base prior are not altered qualitatively by this range of prior strength.

@@

We considered a version of the Gompertz model in which an autoregressive
parameter was fit to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align*}
In addition to the parameters in the original Gompertz model, we estimate an additional parameter $\phi$, which represents the dependence of subsequent residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. The MCMC chains did not converge for \modelsNoConvergeAROne\ populations according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains. This included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr($\nu < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model. We did not include these models in Fig.~\ref{fig:alt}.

@@

Our results indicate that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (TODO FIG).
The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness. Fitting a model with correctly specified observation error did not make substantial improvements to model bias.

%(Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows). Likewise, the other Gompertz parameters were estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols). 
%, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence with observation error is a known phenomenon \citep{knape2012}. Fitting a model with correctly specified observation error made marginal improvements to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).

@@

\singlespacing
\begin{footnotesize}
\begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.3cm}}
\caption{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed Gompertz population dynamics model. Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses. Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events. (TODO add what probability these are.)}\\
\toprule
\input{../analysis/cause-table.tex}
\label{tab:causes-supp}
\end{longtable}
\end{footnotesize}
\onehalfspacing

@@

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/priors-gomp-base.pdf}
\caption{
Probability density of the Bayesian priors for the Gompertz models. From left to right and then top to bottom: (1) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$. The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$. The inset panel shows the same data but with a log-transformed x axis. Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ that we are concerned with.
}
\label{fig:priors}
\end{center}
\end{figure}

@@

@@
